---
title: "Template for Project 2 Data Analysis Report (replace with own title)"
author: "(Your name, student ID here, and your team ID)"
date: "(Date on which the report is written)"
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r, echo=FALSE}
library(tidyverse)
```

# Abstract 

<span style='color:blue'> 
A (very) short summary of the report. As an example, you can simply write one sentence for each section in the report. 
</span>



# Introduction

COVID-19 came to shock the world socially, politically, and economically. Many governments made the decision to lockdown their countries in response to COVID-19 as a way to limit the spread of the virus. This came with direct costs to economic activities as many businesses were forced to halt their operations or adapt to regulations in order to keep productivity afloat. In this paper, we measure the impact of COVID-19 on the economy by examining the effects of COVID-related deaths and cases on GDP. We use GDP as a measurement of the economy as it tracks income and spending in the economy, and it can be compared with other coutries and across time. We also note the various lockdown announcements in March of 2020 to carefully observe the changes in GDP around that time. 
 
# Background 

For our data, we use a csv file containing COVID-19 from the WHO website which can be found [here](https://covid19.who.int/WHO-COVID-19-global-data.csv). This dataset consists of 8 variables: the date reported, country code, country, WHO region, new cases, cumulative cases, new deaths, and cumulative deaths. Information about WHO regions and country codes can be found on the WHO website. The dataset is maintained and updated regularly by WHO. The WHO makes a statement about the dataset in their overview that "all counts are subject to variations in case detection, definitions, laboratory test, and reporting strategies between countries, states, and territories." The case/death data is based on WHO case definitions which can be found [here](https://www.who.int/publications/i/item/who-2019-nCoV-surveillanceguidance-2020.8). The collection of the data was started on the 3rd of January 2020, and collection is still on-going at the time this paper was written. The WHO sources their data through cases and deaths communicated by the International Health Regulation and reconciles this data through official minitries of health websites and social media accounts. Negative numbers may appear in the dataset as reconciliation exercises are performed to keep the data as accurate as possible. 

# Descriptive analysis 

```{r}
covid = read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
covid$Case_mortality_rate = covid$Cumulative_cases / covid$Cumulative_deaths
covid$Case_mortality_rate[is.nan(covid$Case_mortality_rate)] = 0

summary(covid$Date_reported)
summary(covid$New_cases)
summary(covid$Cumulative_cases)
summary(covid$New_deaths)
summary(covid$Cumulative_deaths)
```

# Inferential analysis 

# Model Diagnostics

# Conclusion

<span style='color:blue'> 
Conclude your analysis in this section. You can touch on the following topics. 
</span> 

# Acknowledgement {-}

<span style='color:blue'>
By default, it is assumed that you have discussed this project with your teammates and instructors. List any other people that you have discussed this project with. 
</span>

# Reference {-}

<span style='color:blue'>
List any references you cited in the report. See [here](https://owl.purdue.edu/owl/research_and_citation/apa_style/apa_formatting_and_style_guide/in_text_citations_the_basics.html) for the APA format. 
</span> 

# Session info {-}

<span style='color:blue'>
Report information of your `R` session for reproducibility. 
</span> 


```{r}
sessionInfo()
```