---
title: "Template for Project 2 Data Analysis Report (replace with own title)"
author: "(Your name, student ID here, and your team ID)"
date: "(Date on which the report is written)"
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

***

Note: 

- This template provides only a suggestion. You may use other templates or formats that you are comfortable with. 

- **Remove these remarks in your submission.**

***

# Abstract 

<span style='color:blue'> 
A (very) short summary of the report. As an example, you can simply write one sentence for each section in the report. 
</span>



# Introduction

<span style='color:blue'> 
In this section, state the questions of interest, motivation of this analysis, and potential impact of your results. You can simply rephrase the Project Description for minimal efforts. You can also cite published papers or credible articles on the Internet.
</span>
 
# Background 

<span style='color:blue'> 
In this section, explain the source of data, target population, sampling mechanism, and variables in this data set. You can briefly review  existing research or known results, which will help you in the analysis.
</span> 

# Descriptive analysis 

<span style='color:blue'> 
Select the variables you find relevant based on your understanding in the Background section. Summarize univariate descriptive statistics for the selected variables (mean, standard deviations, missing values,  quantiles, etc.). You can create the table using functions in base `R`, or use packages (see, e.g., [this note](https://cran.r-project.org/web/packages/qwraps2/vignettes/summary-statistics.html)). 
</span>

# Inferential analysis 

# Sensitivity analysis 

# (Optional) Causal interpretation 

# Discussion 

<span style='color:blue'> 
Conclude your analysis in this section. You can touch on the following topics. 
</span> 

# Acknowledgement {-}

<span style='color:blue'>
By default, it is assumed that you have discussed this project with your teammates and instructors. List any other people that you have discussed this project with. 
</span>

# Reference {-}

<span style='color:blue'>
List any references you cited in the report. See [here](https://owl.purdue.edu/owl/research_and_citation/apa_style/apa_formatting_and_style_guide/in_text_citations_the_basics.html) for the APA format. 
</span> 

# Session info {-}

<span style='color:blue'>
Report information of your `R` session for reproducibility. 
</span> 


```{r}
sessionInfo()
```